---
title: "Benchmark Results: GPyTorch Learner"
author: "Khalifa Seck"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmark Results: GPyTorch Learner}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

Gaussian Process regression learner for mlr3 using GPyTorch with GPU acceleration support.

## Installation

### Complete Installation (Recommended)

```{r installation}
# Step 1: Install R package
remotes::install_github("KhalifaSeck/mlr3learners.gpytorch")

# Step 2: Install miniconda (one-time only)
if(!reticulate::miniconda_exists()) {
  reticulate::install_miniconda()
}

# Step 3: Create Python environment (one-time only)
if(!reticulate::conda_exists("gpytorch")) {
  reticulate::conda_create("gpytorch")
  reticulate::conda_install("gpytorch", c("pytorch", "gpytorch"))
}

# Step 4: Configure Python path
Sys.setenv(RETICULATE_PYTHON="~/AppData/Local/r-miniconda/envs/gpytorch/python.exe")  # Windows
# Sys.setenv(RETICULATE_PYTHON="~/miniconda3/envs/gpytorch/bin/python")  # Linux/Mac

# Step 5: Activate environment
reticulate::use_condaenv("gpytorch", required = TRUE)
```

### Verify Installation

```{r verify}
library(mlr3)
library(mlr3learners.gpytorch)

task <- mlr3::tsk("mtcars")
learner <- mlr3::lrn("regr.gpytorch", n_iter = 50)
learner$train(task)
pred <- learner$predict(task)
print(pred)
```

## Usage

### Basic Example

```{r usage}
library(mlr3)
library(mlr3learners.gpytorch)
library(reticulate)

use_condaenv("gpytorch")

task <- tsk("mtcars")
learner <- lrn("regr.gpytorch", n_iter = 50)

learner$train(task)
pred <- learner$predict(task)
print(pred)
```

### GPU Usage

```{r gpu_usage}
learner_gpu <- lrn("regr.gpytorch", n_iter = 50, device = "cuda")
learner_cpu <- lrn("regr.gpytorch", n_iter = 50, device = "cpu")
learner_auto <- lrn("regr.gpytorch", n_iter = 50, device = "auto")  # Auto-detects GPU
```

## Hyperparameters

| Parameter | Type | Default | Range | Description |
|-----------|------|---------|-------|-------------|
| kernel | factor | "rbf" | {rbf, matern} | Kernel type for covariance function |
| lr | numeric | 0.1 | [0.001, 1] | Learning rate for Adam optimizer |
| n_iter | integer | 50 | [10, 500] | Number of training iterations |
| device | factor | "auto" | {auto, cpu, cuda} | Computing device (auto detects GPU) |

## Benchmark Results

Performance evaluated using Mean Squared Error (MSE) in 5-fold cross-validation.

### Benchmark Code

```{r benchmark_code}
library(mlr3verse)
library(ggplot2)

# Setup
use_condaenv("gpytorch")

# Tasks
task_air = as_task_regr(na.omit(airquality), target = "Ozone", id = "airquality")
task_boston = as_task_regr(MASS::Boston, target = "medv", id = "boston")

# Learners
learners = list(
  lrn("regr.gpytorch", n_iter = 50),
  lrn("regr.kknn"),
  lrn("regr.cv_glmnet"),
  lrn("regr.featureless")
)

# Benchmark
design = benchmark_grid(
  tasks = list(task_air, task_boston),
  learners = learners,
  resamplings = rsmp("cv", folds = 5)
)

bmr = benchmark(design)
results = bmr$aggregate(msr("regr.mse"))
print(results)

# Plot
ggplot(bmr$score(msr("regr.mse")), aes(x = regr.mse, y = learner_id)) +
  geom_point(size = 3, alpha = 0.7) +
  facet_grid(task_id ~ ., scales = "free_x") +
  labs(x = "Mean Squared Error (MSE)", y = "Algorithm",
       title = "Prediction Error in 5-Fold Cross-Validation") +
  theme_bw()
```

### Results

| Dataset | GPyTorch | KNN | CV-Glmnet | Featureless |
|---------|----------|-----|-----------|-------------|
| airquality | 398.58 | 388.28 | 575.86 | 1107.45 |
| boston | **9.68** | 18.02 | 27.79 | 84.70 |

Lower MSE indicates better performance. GPyTorch ranks #1 on boston dataset.

### Algorithm Rankings

**airquality dataset (111 observations)**

1. KNN: 388.28 MSE
2. GPyTorch: 398.58 MSE
3. CV-Glmnet: 575.86 MSE
4. Featureless: 1107.45 MSE

**boston dataset (506 observations)**

1. **GPyTorch: 9.68 MSE** 
2. KNN: 18.02 MSE
3. CV-Glmnet: 27.79 MSE
4. Featureless: 84.70 MSE

### Analysis

**Does GPyTorch learn non-trivial patterns?**

- airquality: 64% improvement over featureless baseline
- boston: 89% improvement over featureless baseline

 Yes, GPyTorch successfully learns non-trivial patterns on both datasets.

**Is GPyTorch competitive with other algorithms?**

- Ranks #1 on boston dataset
- Ranks #2 on airquality dataset
- Outperforms linear models on both datasets
- Competitive with KNN

 GPyTorch demonstrates excellent performance, particularly on larger datasets.

## GPU vs CPU Performance

Performance comparison on NVIDIA GeForce RTX 3050 Laptop GPU.

### Prediction Accuracy

GPU and CPU produce identical predictions:

- airquality: 398.58 (GPU) = 398.58 (CPU)
- boston: 9.68 (GPU) = 9.68 (CPU)

 Confirms correct implementation across both backends.

### Training Time (30 iterations)

**airquality (111 obs):** CPU 2.5x faster (GPU overhead dominates)  
**boston (506 obs):** GPU 1.3x faster (computation outweighs overhead)

**Recommendation:** Use CPU for small datasets (<200 obs), GPU for large datasets (>1000 obs).

## Technical Features

- **GPU/CPU support**: Automatic device detection or manual selection
- **Feature normalization**: Automatic zero mean, unit variance scaling
- **Kernel options**: RBF and Matérn kernels
- **Numerical stability**: Proper variance handling

## System Requirements

**CPU:** Modern multi-core CPU, 4GB RAM  
**GPU:** NVIDIA GPU with CUDA 11.8+, 4GB GPU memory

## Citation

```bibtex
@inproceedings{gardner2018gpytorch,
  title={GPyTorch: Blackbox Matrix-Matrix Gaussian Process Inference with GPU Acceleration},
  author={Gardner, Jacob R and Pleiss, Geoff and Weinberger, Kilian Q and Bindel, David and Wilson, Andrew Gordon},
  booktitle={Advances in Neural Information Processing Systems},
  year={2018}
}
```

## Author

Khalifa Seck - [GitHub](https://github.com/KhalifaSeck)

## License

MIT License
