---
title: "Benchmark Results: GPyTorch Learner"
author: "Khalifa Seck"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmark Results: GPyTorch Learner}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

This vignette presents comprehensive benchmark results for the mlr3learners.gpytorch package, comparing GPyTorch Gaussian Process regression against other standard algorithms.

## Installation

```{r installation}
remotes::install_github("KhalifaSeck/mlr3learners.gpytorch")
```

## Datasets

Two regression datasets were used for evaluation:

- **airquality**: 111 observations, predicting Ozone levels
- **boston**: 506 observations, predicting median house values

## Benchmark Setup

5-fold cross-validation with the following learners:

1. GPyTorch (CPU)
2. GPyTorch (GPU)
3. K-Nearest Neighbors (auto-tuned k=1-30)
4. CV-Glmnet (penalized linear regression)
5. Featureless baseline

## Results

### Overall Performance (Mean MSE)

| Dataset | GPyTorch (GPU) | GPyTorch (CPU) | KNN | CV-Glmnet | Featureless |
|---------|----------------|----------------|-----|-----------|-------------|
| airquality | 398.58 | 398.58 | 388.28 | 575.86 | 1107.45 |
| boston | 9.68 | 9.68 | 18.02 | 27.79 | 84.70 |

### Rankings

**airquality dataset:**

1. KNN: 388.28 MSE
2. GPyTorch: 398.58 MSE (64% improvement over baseline)
3. CV-Glmnet: 575.86 MSE
4. Featureless: 1107.45 MSE

**boston dataset:**

1. GPyTorch: 9.68 MSE (89% improvement over baseline)
2. KNN: 18.02 MSE
3. CV-Glmnet: 27.79 MSE
4. Featureless: 84.70 MSE

## GPU vs CPU Performance

Tested on NVIDIA GeForce RTX 3050 Laptop GPU.

### Training Time (30 iterations)

**airquality (111 observations):**

- CPU: 0.19-0.24 seconds
- GPU: 0.31-0.48 seconds
- Result: CPU 2x faster (GPU overhead dominates)

**boston (506 observations):**

- CPU: 0.34-0.37 seconds
- GPU: 0.28-0.33 seconds
- Result: GPU 1.3x faster

### Analysis

GPU overhead (memory transfer) is approximately 0.2 seconds. For small datasets, this overhead exceeds computational gains. For larger datasets (500+ observations), GPU acceleration becomes beneficial.

## Conclusion

GPyTorch demonstrates:

- Excellent learning capability (64-89% improvement over baseline)
- Competitive performance (rank 1 on boston, rank 2 on airquality)
- GPU acceleration beneficial for datasets with 500+ observations
- Identical predictions on CPU and GPU (validates implementation)

See the main [README](https://github.com/KhalifaSeck/mlr3learners.gpytorch) for complete details.